Predicting Diabetes Based on Various Factors Using Decision Trees

### 2023-12-05 manuscript/manuscript.tex
Double space pdf
Moving of roadmap
Table of head of dataset
Movement of paragraph in method section
Change to decision tree

\documentclass[12pt]{article}

@@

\title{Predicting Diabetes Based on Various Factors Using logistic Regression}
\author{Ashley Merritt\\
    Department of Statistics\\
    University of Connecticut
    }

@@

@@

    Previously, the prediction of diabetes has been studied using machine learning models. Specifically in Sisodia's paper, it focused
    on the "Prediction of Diabetes using Classification Algorithms", they worked through the topic by producing results using support 
    vector machine, Naive Bayes classifier, and decision tree classifier (\cite{Sisodia2018Prediction}). In this paper, I will be 
    continuing the investigation with a different data set using logistic regression and various other statistical tests.

\section{Data}
\label{sec:data}
    In order to study the proposed research question on the topic of diabetes, I searched through many databases in order to find a set 
    that had all of the information I was looking for. After exploration, I found a data set entitled "Healthcare Diabetes Data set" on 
    the website Kaggle. The data set was originally sourced from the National Institute of Diabetes and Digestive and Kidney Diseases. The 
    data encompasses eight different predictors including pregnancies, glucose, blood pressure, skin thickness, insulin, BMI, diabetes 
    pedigree function, and age.

@@

\section{Methods}
\label{sec:meth}
    Various methods will be used to complete the statistical analysis necessary for this project. I will begin by exploring basic 
    descriptive statistics to gain an understanding of my data set. I will then progress to creating visualizations including histograms 
    and box plots to understand relationships and patterns within my data. In order to draw insight on my data I will perform logistic 
    regression analysis to determine which predictors are significant in predicting if someone has diabetes. In order to perform the 
    logistic regression, I will have to check the necessary assumptions associated with the method. Within this analysis I will perform 
    necessary transformations in order to create the best model that I can. From my results I will draw necessary conclusions about my data.

    The first assumption that will be checked in that the dependent variable is binary, this true as our outcome variable is in the form of 
    1 (having diabetes) or 2 (not having diabetes). The next assumption that will be checked is if the observations are independent of each 
    other, this is also true for our data as each predictor is calculated or taken independent of one another. The third assumption that 
    will be checked is that there is no multicollinearity among the independent variables. In order to check this assumption we will 
    calculate the variance inflation factor, VIF, in order to measure the strength of the correlation between predictor variables in our 
    model. One example of this is seen in Table~\ref{tab:vif} where all of the values are around 1, indicating that there is moderate 
    correlation between pregnancies and the other predictors but this value is not in a concerning range.

\begin{table}[ht]
\caption{VIF Values} 
\label{tab:vif}
\centering
\begin{tabular}{rr}
  \hline
 & vif(lm(P \~{} G + BP + ST + I + BMI + DPF + A)) \\ 
  \hline
Glucose & 1.29 \\ 
  BloodPressure & 1.18 \\ 
  SkinThickness & 1.50 \\ 
  Insulin & 1.44 \\ 
  BMI & 1.30 \\ 
  DiabetesPedigreeFunction & 1.06 \\ 
  Age & 1.18 \\ 
   \hline
\end{tabular}
\end{table}

    After exploring the descriptive statistics, I will begin to perform a logistic regression. Logistic regression, is used to describe the 
    data and the relationship between one dependent variable against many independent variables. In this particular case, we will be looking 
    at the binary response of diabetes with the various independent variables including but not limited to; pregnancies and glucose levels. 
    To begin, I will split the data into a training and a test set. This will be done by setting a random seed of 30 and using 80 percent 
    of the data set in the training set and the remaining 20 percent as the testing set. We will next begin to build the logistic 
    regression model using the training data.

    In Table~\ref{tab:sl} you will see the output of each predictor on predicting the outcome of diabetes. This table will provide us with 
    coefficients, the standard errors, the z-scores, and finally the p-value for each predictor in the model. Based on the output, we can 
    determine that only skin thickness is not statistically significant in predicting if a patient may have diabetes. Therefore, all of the 
    other predictors are statistically significant and should be monitored in predicting diabetes.

\begin{table}[ht]
    \caption{Summary of Logistic Regression Model}
    \label{tab:sl}
\centering
\begin{tabular}{lrrrr}
  \hline
Predictor & Estimate & Std. Error & z value & Pr($>$$|$z$|$) \\ 
  \hline
(Intercept) & -8.02 & 0.41 & -19.73 & 0.00 \\ 
  Pregnancies & 0.11 & 0.02 & 6.04 & 0.00 \\ 
  Glucose & 0.03 & 0.00 & 16.34 & 0.00 \\ 
  BloodPressure & -0.01 & 0.00 & -3.69 & 0.00 \\ 
  SkinThickness & 0.00 & 0.00 & 0.26 & 0.79 \\ 
  Insulin & -0.00 & 0.00 & -2.74 & 0.01 \\ 
  BMI & 0.08 & 0.01 & 9.07 & 0.00 \\ 
  DiabetesPedigreeFunction & 0.76 & 0.17 & 4.38 & 0.00 \\ 
  Age & 0.02 & 0.01 & 3.03 & 0.00 \\ 
   \hline
\end{tabular}
\end{table}
\section{Results}
\label{sec:resu}

>>>>

\documentclass[12pt]{article}
\usepackage{setspace}
\doublespacing

@@

\title{Predicting Diabetes Based on Various Factors Using Decision Trees}
\author{Ashley Merritt\\
    Department of Statistics\\
    University of Connecticut
    }

@@

    Previously, the prediction of diabetes has been studied using machine learning models. Specifically in Sisodia's paper, it focused
    on the "Prediction of Diabetes using Classification Algorithms", they worked through the topic by producing results using support 
    vector machine, Naive Bayes classifier, and decision tree classifier (\cite{Sisodia2018Prediction}). In this paper, I will be 
    continuing the investigation with a different data set using decision trees.

@@

\section{Data}
\label{sec:data}
    In order to study the proposed research question on the topic of diabetes, I searched through many databases in order to find a set 
    that had all of the information I was looking for. After exploration, I found a data set entitled "Healthcare Diabetes Data set" on 
    the website Kaggle. The data set was originally sourced from the National Institute of Diabetes and Digestive and Kidney Diseases. The 
    data encompasses eight different predictors including pregnancies, glucose, blood pressure, skin thickness, insulin, BMI, diabetes 
    pedigree function, and age. In Table~\ref{tab:head_of_data} I will provide the first couple lines of dataset in order to provide
    more information on the dataset.

    \begin{table}[ht]
    \caption{Head of the Dataset}
    \centering
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{rrrrrrrrr}
      \hline
    Pregnancies & Glucose & BloodPressure & SkinThickness & Insulin & BMI & DiabetesPedigreeFunction & Age & Outcome \\ 
      \hline
      6 & 148 &  72 &  35 &   0 & 33.60 & 0.63 &  50 &   1 \\ 
        1 &  85 &  66 &  29 &   0 & 26.60 & 0.35 &  31 &   0 \\ 
        8 & 183 &  64 &   0 &   0 & 23.30 & 0.67 &  32 &   1 \\ 
        1 &  89 &  66 &  23 &  94 & 28.10 & 0.17 &  21 &   0 \\ 
        0 & 137 &  40 &  35 & 168 & 43.10 & 2.29 &  33 &   1 \\ 
        5 & 116 &  74 &   0 &   0 & 25.60 & 0.20 &  30 &   0 \\ 
       \hline
    \end{tabular}%
    } 
    \label{tab:head_of_data}
    \end{table}

@@

\section{Methods}
\label{sec:meth}
    Various methods will be used to complete the statistical analysis necessary for this project. I will begin by exploring basic 
    descriptive statistics to gain an understanding of my data set. I will then begin to create my model for my decision tree by
    splitting the data into a test and training set. The training set will encompass 80 percent of the data whereas the test set will
    only have 20 percent. After creating these two sets we will now be ready to build the model, this will be done by using the rpart
    function in R, where method will be 'class' as this is a binary model. We will then see in Figure~\ref{tab:structure} the structure
    of the decision tree created. Looking at the leftmost path in the decision tree we can see that at the top the outcome was 0 meaning 
    that the patient did not have diabetes 35 percent of the time. Moving down to the next node, we can see that 68 percent of patients with
    glucose level less than 128, did not have diabetes 19 percent of the time. To finish this path, 61 percent of the time if the patient was
    older than 29 years old, then 8 percent of the time the patient had diabetes. The rest of the structure is similar but with different 
    values and predictors.

\begin{figure}[tbp]
  \centering
  \caption{Decision Tree Structure}
  \includegraphics[width=\textwidth]{Decision Tree Structure.pdf}
  \label{fig:structure}
  \end{figure}

\section{Results}
\label{sec:resu}
